---
title: "X chromosome signatures"
author: "Rachael Caelie (Rocky) Aikens"
date: "12/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
knitr::opts_knit$set(root.dir = '../../data')

library(ggplot2)
library(gplots)
library(stats)
library(readr)
library(dplyr)
library(knitr)
require(reshape2)
library(gridExtra)
library(grid)
library(DescTools)
require("ggrepel")
theme_set(theme_light())
```

# Introduction

Because of the nature of sex chromosome inheritance and expression, patterns of polymorphism and mutation rate may differ on the X Chromosome compared to the autosome.  For this reason, we wanted to understand whether what signatures of variation in polymorphism appear when we restrict our analysis to the X Chromosome.

## A Cross Continental Test for Heterogeneity in 3-mer Polymorphism Counts on the X Chromosome 

```{r T1 setup methods}
# calculates homogeneity test p values for Fourway comparisons of counts dfs
# currently only supports 3mer dataframes when calculating sums of autosomal variants
XChrom.BD <- function(AFR, EUR, EAS, SAS, filter = T){
  # add autosomal columns to result
  AFR$Auto <- rowSums(AFR[,6:27])
  EUR$Auto <- rowSums(EUR[,6:27])
  EAS$Auto <- rowSums(EAS[,6:27])
  SAS$Auto <- rowSums(SAS[,6:27])
  
  n.contexts = length(AFR$Context)
  
  # make dataframe for results
  result <- data.frame(matrix(ncol=15,nrow=n.contexts))
  colnames(result) <- c("Context", "X5mer","X3mer", "X1mer", 
                        "AFR.Auto", "AFR.X", "EUR.Auto", "EUR.X", 
                        "EAS.Auto", "EAS.X", "SAS.Auto", "SAS.X", "BDp", "p", "CommonOR")
  result$Context <- AFR$Context
  result$X5mer <- AFR$X5mer # for smaller contexts, X3mer and X5mer columns do not exist,
  result$X3mer <- AFR$X3mer # and will disappear at this step
  result$X1mer <- AFR$X1mer 
  result$AFR.Auto <- AFR$Auto; result$EUR.Auto <- EUR$Auto
  result$EAS.Auto <- EAS$Auto; result$SAS.Auto <- SAS$Auto
  result$AFR.X <- AFR$chrX; result$EUR.X <- EUR$chrX
  result$EAS.X <- EAS$chrX; result$SAS.X <- SAS$chrX
  
  # start setting up tables
  sums.Auto <- c(sum(AFR$Auto), sum(EUR$Auto), sum(EAS$Auto), sum(SAS$Auto))
  sums.X <- c(sum(AFR$chrX), sum(EUR$chrX), sum(EAS$chrX), sum(SAS$chrX))
  
  # set up table and run test for each context
  for (i in 1:n.contexts){
    mut.Auto <- c(AFR$Auto[i], EUR$Auto[i], EAS$Auto[i], SAS$Auto[i])
    mut.X <- c(AFR$chrX[i], EUR$chrX[i], EAS$chrX[i], SAS$chrX[i])
    nmut.Auto <- sums.Auto - mut.X
    nmut.X <- sums.X - mut.X
    data <- array(rep(1, 2*2*4), dim=c(2, 2, 4))
    
    data[2,1,] <- mut.X
    data[2,2,] <- nmut.X
    data[1,1,] <- mut.Auto
    data[1,2,] <- nmut.Auto 
    
    result$BDp[i] <- BreslowDayTest(data)$p.value
    result$p[i] <- mantelhaen.test(data)$p.value
    result$CommonOR[i] <- as.numeric(mantelhaen.test(data)$estimate)
  }
    
  return(result)
}

fdr <- function(p.data){
  p.data <- p.data[complete.cases(p.data),]
  
  # This uses Benjamini-Hochberg-Yekutieli fdr
  p.data$fdr <- p.adjust(p.data$p, method = "fdr")
  
  # multiple hypothesis correction by holm
  p.data$holm <- p.adjust(p.data$p, method = "holm")
  
  alpha = 0.05/length(p.data$p)
  p.data <- p.data[complete.cases(p.data), ]
  
  n.sig <- c(length(p.data$p), sum(p.data$p < alpha), sum(p.data$holm< 0.05), 
             sum(p.data$fdr< 0.1), sum(p.data$fdr< 0.05), 
             sum(p.data$fdr< 0.01), sum(p.data$fdr< 0.001))
  
  names(n.sig) <- c("Total tests","Bonferroni", "Holm", 
                    "FDR<0.1", "FDR<0.05", 
                    "FDR<0.01", "FDR<0.001")
  
  return(list(n.sig, p.data))
}
```

```{r 1kg setup data}
AFR_3mer_counts <- read_delim("3mer/AFR_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
  
EUR_3mer_counts <- read_delim("3mer/EUR_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)

EAS_3mer_counts <- read_delim("3mer/EAS_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE) 

SAS_3mer_counts <- read_delim("3mer/SAS_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE) 

```

```{r}
p.X <- XChrom.BD(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts)

fdr.P.X <- fdr(p.X)

write.table(x = fdr.P.X[[2]], file = "../workflows_finished/X_Chromosome_only/3mer_Xtest.txt", quote = F, sep = "\t", row.names = F)
```

The following are results from a Breslow Day and CMH test for each 3-mer polymorphism type.

```{r}
print(fdr.P.X)
```

