---
title: "Supplementary Notes"
author: "Rachael C. Aikens, Benjamin F. Voight"
date: "March 27, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
knitr::opts_knit$set(root.dir = '../../data')

library(ggplot2);library(gplots)
library(stats);library(readr)
library(dplyr);library(knitr)
require(reshape2)
library(gridExtra);library(grid)
require(ggrepel)
```

# Supplementary Figures

```{r S1 upload data}
AFR_3mer_counts <- read_delim("3mer/AFR_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
EUR_3mer_counts <- read_delim("3mer/EUR_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
EAS_3mer_counts <- read_delim("3mer/EAS_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
SAS_3mer_counts <- read_delim("3mer/SAS_3mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
```

```{r S1 functions}
# calculates homogeneity test p values for pairwise comparisons of two dfs of counts
pairwise.chi <- function(counts.1, counts.2, filter = T){
  n.contexts = length(counts.1$Context)
  result <- data.frame(matrix(ncol=4,nrow=n.contexts))
  colnames(result) <- c("Context", "Counts1", "Counts2", "p")
  
  result$Context <- counts.1$Context
  sum.1 <- sum(counts.1$Count)
  sum.2 <- sum(counts.2$Count)

  for (i in 1:n.contexts){
    c.a <- c(counts.1$Count[i], counts.2$Count[i])
    c.b <- c(sum.1, sum.2) - c.a
    data <- cbind(c.a, c.b)
    
    result$Counts1[i] <- c.a[1]
    result$Counts2[i] <- c.a[2]
    warning <- is(tryCatch(chisq.test(data), warning = function(w) w), "warning")
    if (filter == T & warning){
      result$p[i] <- NA
    }
    else result$p[i] <- chisq.test(data)$p.value
  }
  
  return(result)
}

#color vector so that plot colorings match Harris
h.colors <- c("dark blue", "magenta", "purple", "forest green", "#0099FF", "red")

#volcano plot function as in Harris 2015
volcano.plot <- function(counts.1, counts.2, 
                         p.vals, lab.lim, 
                         pops = c("Population 1", "Population 2"), 
                         legpos = c(0.2, 0.75)){
  
  f.1 <- counts.1$Count/sum(counts.1$Count)
  f.2 <- counts.2$Count/sum(counts.2$Count)
  f.diff <- (f.1 -f.2)/f.2
  alpha <- 0.05/(6*length(counts.1$Count))
  
  sigs <- subset(counts.1, p.vals$p < lab.lim)
  f.diff.sigs <- subset(f.diff, p.vals$p < lab.lim)
  p.sigs <- subset(p.vals, p.vals$p < lab.lim)
  
  v_plot <- qplot(f.diff, -log10(p.vals$p)) +
    labs(x = paste("Fold excess in", pops[1]) , 
         y = expression(-log[10](italic(p))), 
         title = paste(pops[1], "v", pops[2])) +
    scale_color_manual("", values = h.colors)+ # set colors
    geom_text(data = sigs, aes(f.diff.sigs, -log10(p.sigs$p), label = sigs$Context, color = sigs$X1mer), 
              vjust = 1, nudge_y =-1, hjust = 1, nudge_x = -.03, size = 3, check_overlap = TRUE)+ # label highly significant
    geom_point(aes(color=factor(counts.1$X1mer)), size = 3)+ # Color pts by one_mer
    geom_hline(yintercept = -log10(alpha), color = "black", linetype = 2)+# significance line
    theme(axis.text.x = element_text(size = rel(1.3)), axis.title.x = element_text(size = rel(1.1)), # adjust text sizes
          axis.title.y = element_text(size = rel(1.1)), axis.text.y = element_text(size = rel(1.35)), 
          legend.text = element_text(size = rel(1.1)), legend.position = legpos,
          title = element_text(size = rel(1.2))) + # legend position and size 
    xlim(c(-.4, .6)) + ylim(c(0, 400))
  
  return(v_plot)
}
```

```{r S1 make figure}
pEURvAFR <- pairwise.chi(EUR_3mer_counts, AFR_3mer_counts, filter = F)
a <- volcano.plot(EUR_3mer_counts, AFR_3mer_counts, pEURvAFR, 1e-25, c("Europe", "Africa"), legpos = "none")
pEASvAFR <- pairwise.chi(EAS_3mer_counts, AFR_3mer_counts, filter = F)
b <- volcano.plot(EAS_3mer_counts, AFR_3mer_counts, pEASvAFR, 1e-25, c("East Asia", "Africa"))
grid.arrange(a,b, ncol= 2)
```

**Supplementary Figure 1:** *Replication of Figure 1 from Kelley Harris, 2015*. Using our variant filtration pipleline for extracting population-private variants from the Phase III 1,000 Genomes Dataset, we recapitulate Harris's previous findings from analysis of the Phase I Release.  In keeping with Harris, 2015, the proportions of private 3mer substitutions were compared between pairs of populations using a pairwise chi-squared test.  Ordered p-value correction was not applied.  Unlike Harris, we choose to consider reverse-complimentary substitution classes as identical (e.g. TCC$\rightarrow$TTC and GGA$\rightarrow$ GAA are considered equivalent).

***********************************************************************************************************

\pagebreak

```{r S2 upload data}
# Africa
AFR_AF_bins <- read_delim("by_AF/v1_m_4000/AFR_sortedSNPs_bins.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
AFR_AF_counts <- read_delim("by_AF/v1_m_4000/AFR_AF_3mer_counts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)

# Europe
EUR_AF_bins <- read_delim("by_AF/v1_m_4000/EUR_sortedSNPs_bins.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
EUR_AF_counts <- read_delim("by_AF/v1_m_4000/EUR_AF_3mer_counts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)

# East Asia
EAS_AF_bins <- read_delim("by_AF/v1_m_4000/EAS_sortedSNPs_bins.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
EAS_AF_counts <- read_delim("by_AF/v1_m_4000/EAS_AF_3mer_counts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)

# South Asia
SAS_AF_bins <- read_delim("by_AF/v1_m_4000/SAS_sortedSNPs_bins.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
SAS_AF_counts <- read_delim("by_AF/v1_m_4000/SAS_AF_3mer_counts.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
```

```{r S2 functions}
SFS.plot <- function(AFR, AFR.bins, EUR, EUR.bins, EAS, EAS.bins, SAS, SAS.bins, mut, before, leg = T, xlabel = T){
  i <- which(AFR$Context == mut)
  n <- nchar(mut)
  ref <- substr(mut, 1, n-3)
  alt <- substr(mut, n, n)
  
  #based on sequence context, figure out which text columns to remove
  remove <- c(seq(before+1))
  
  #calculate mutation proportion across AF bin for each pop
  AFR_prop <- AFR[i, -remove]/colSums(AFR[,-remove])
  EUR_prop <- EUR[i, -remove]/colSums(EUR[,-remove])
  EAS_prop <- EAS[i, -remove]/colSums(EAS[,-remove])
  SAS_prop <- SAS[i, -remove]/colSums(SAS[,-remove])
  
  props <- as.numeric(c(AFR_prop, EUR_prop, EAS_prop, SAS_prop))
  bins <- c(AFR.bins$AVG_AF, EUR.bins$AVG_AF, EAS.bins$AVG_AF, SAS.bins$AVG_AF)
  
  labels <- c(rep("Africa", length(AFR_prop)), rep("Europe", length(EUR_prop)), 
              rep("East Asia", length(EAS_prop)), rep("South Asia", length(SAS_prop)))
  dat <- data.frame(cbind(props, bins, labels))
  names(dat)<- c("proportions", "AF", "POP")
  
  YU = max(props)
  YL = min(props)
  XU = max(log10(bins))
  XL = min(log10(bins))
  
  #ggplot method
  SFSplot <- ggplot(dat, aes(x = bins, y = props, group = labels, color = labels),
                    xlim = c(XL, XU), ylim = c(YL, YU)) +
              geom_line(size = 1)+
              scale_y_continuous(name = bquote(.(ref)%->%.(alt) ~ .("proportion"))) +
              scale_x_log10(name = "Alternate allele frequency") +
              scale_color_manual("", values = c("forestgreen", "red", "darkblue", "magenta"))
  
  if (!leg){
    SFSplot <- SFSplot + theme(legend.position = "none")
  }
  else {
    SFSplot <- SFSplot + theme(legend.position = c(0.16, 0.3), 
                               legend.title = element_blank(), 
                               legend.key.size = unit(0.5, "cm"),
                               legend.margin = unit(0.7, "cm"))
  }
  
  if (!xlabel){
    SFSplot <- SFSplot + theme(axis.text.x = element_blank(),
                               axis.title.x = element_blank(),
                               axis.ticks.x = element_blank())
  }

  return(SFSplot)
}
```

```{r S2 make figure, fig.height= 4}
a <- SFS.plot(AFR_AF_counts, AFR_AF_bins, EUR_AF_counts, EUR_AF_bins, 
         EAS_AF_counts, EAS_AF_bins, SAS_AF_counts, SAS_AF_bins, "TCG->T", 1, leg = T, xlabel = F)
b <- SFS.plot(AFR_AF_counts, AFR_AF_bins, EUR_AF_counts, EUR_AF_bins, 
         EAS_AF_counts, EAS_AF_bins, SAS_AF_counts, SAS_AF_bins, "ACG->T", 1, leg = F, xlabel = F)
c <- SFS.plot(AFR_AF_counts, AFR_AF_bins, EUR_AF_counts, EUR_AF_bins, 
         EAS_AF_counts, EAS_AF_bins, SAS_AF_counts, SAS_AF_bins, "GCG->T", 1, leg = F)
d <- SFS.plot(AFR_AF_counts, AFR_AF_bins, EUR_AF_counts, EUR_AF_bins, 
         EAS_AF_counts, EAS_AF_bins, SAS_AF_counts, SAS_AF_bins, "CCG->T", 1, leg = F)

grid.arrange(a, b, c, d, ncol = 2)
```

**Supplementary Figure 2:** *No doubleton excess in CpG mutation profile*.  Plots of the proportion of each type of CpG substitution in each population across allele frequency bins.  Since singletons are removed from this analysis, a doubleton excess would be reflected in an especially high proportion of CpG mutations in the lowest alternative allele frequency bin.  While we do not observe a doubleton excess, we do see an unexplained depletion of tripleton CpG variants in Asia and Africa for most CpG substitutions, reflected in the immediate dip in substitution proportion in the second smallest allele frequency bin for African and Asian populations.

***********************************************************************************************************

\pagebreak

```{r S3 upload data}
gw_3mer_counts <- read_delim("gw_counts/gw_3mer_counts.txt", 
                             "\t", escape_double = FALSE, trim_ws = TRUE)
```

```{r S3 functions}
CI.plot <- function(AFR, EUR, EAS, SAS, mut, small = F) {
  n <- nchar(mut)
  ref <- substr(mut, 1, n-3)
  alt <- substr(mut, n, n)
  mut.i <- which(AFR$Context == mut)
  popnames <- c("AFR","EUR", "SAS", "EAS")
  colors <- c("forestgreen", "darkblue", "magenta","red")
  
  #have to do a silly workaround or R will sort popnames alphanumerically
  poplabs <- factor(popnames, levels= popnames)
  
  #cycle through pops and get counts for mut
  counts <- rep(0, 4)
  sums <- rep(0, 4)
  pops <- list(AFR, EUR, SAS, EAS)
  
  for (i in 1:length(pops)){
    counts[i] <- pops[[i]]$Count[mut.i]
    sums[i]<- sum(pops[[i]]$Count)
  }
  
  #estimate substitution probability
  N.c <- AFR$context_in_genome[mut.i] 
  theta <- counts/N.c
  L <- theta - 1.96*sqrt(theta*(1-theta)/N.c)
  U <- theta + 1.96*sqrt(theta*(1-theta)/N.c)
  
  #normalize to rate estimate; assume genome wide subsitution probability is measured without error
  norm <- 1.2E-8*(sum(as.numeric(AFR$context_in_genome))/3)/sums
  
  df <- data.frame(cbind(popnames, theta*norm, L*norm, U*norm))
  plotcol <- reorder(colors, theta*norm)
  
  CIplot <- ggplot(df, aes(reorder(popnames, c(1,2,3,4)), theta*norm)) +
    geom_point(size = 2, color = plotcol) +
    geom_errorbar(aes(ymax = U*norm, ymin = L*norm), color = plotcol, size =.75)+
    labs(title = bquote(.(ref)%->%.(alt) ~ "mutation rate"), y = NULL) + #y axis label
    theme(axis.text.x = element_text(size = rel(.7)), axis.title.x = element_blank(), #adjust text sizes
          axis.title.y = element_text(size = rel(.7)), axis.text.y = element_text(size = rel(.7), angle = 0, hjust = 0.5), title = element_text(size = rel(.5)),
          legend.position = 'none')
  
  if (small) {
    CIplot <- CIplot + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())+
      labs(title = bquote(.(ref)%->%.(alt)))
  }
  
  return(CIplot)
}
```

```{r S3 make figure, fig.height = 8, fig.width=7}
a <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "TCC->T", small = F)
b <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "ACC->T", small = F)
c <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "TCT->T", small = F)
d <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "CCC->T", small = F)
e <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "TCG->T", small = F)
f <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "ACG->T", small = F)
g <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GCG->T", small = F)
h <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GAT->T", small = F)
i <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "ACC->A", small = F)
j <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "ACA->T", small = F)
k <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "TCA->T", small = F)
l <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "ACT->T", small = F)
m <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GCT->T", small = F)
n <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GAC->T", small = F)
o <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GCC->T", small = F)


grid.arrange(a,b,c,d,e,f,g,h,i,j,k,l,m,n,o, ncol = 4)
```

**Supplementary Figure 3:** Approximate 95% confidence interval estimates of inferred mutation rate for each highly significantly variable 3-mer type across Africa (AFR), Europe (EUR), South Asia (SAS), and East Asia (EAS).


***********************************************************************************************************

\pagebreak

```{r S4 functions}
#box plot of rate by chrom for a particular sequence context
chrom.box <- function(AFR, EUR, EAS, SAS, gw, mut){
  #get data
  chrom.dat <- chrom.process.data(AFR, EUR, EAS, SAS, gw, mut)
  
  n <- nchar(mut)
  ref <- substr(mut, 1, n-3)
  alt <- substr(mut, n, n)
  
  #find outliers
  AFR.rates <- subset(chrom.dat, chrom.dat$pop == "AFR")
  AFR.outliers <- subset(AFR.rates, is_outlier(AFR.rates$rate)==TRUE)
  EUR.rates <- subset(chrom.dat, chrom.dat$pop == "EUR")
  EUR.outliers <- subset(EUR.rates, is_outlier(EUR.rates$rate)==TRUE)
  EAS.rates <- subset(chrom.dat, chrom.dat$pop == "EAS")
  EAS.outliers <- subset(EAS.rates, is_outlier(EAS.rates$rate)==TRUE)
  SAS.rates <- subset(chrom.dat, chrom.dat$pop == "SAS")
  SAS.outliers <- subset(SAS.rates, is_outlier(SAS.rates$rate)==TRUE)
  
  #plot
  c_plot <- ggplot(chrom.dat, aes(pop, rate))+
    geom_boxplot(outlier.color = NA, fill = c("palegreen1",  "steelblue1", 'plum1', "lightcoral"))+
    labs(x = "\nPopulation", title = bquote(.(ref)%->%.(alt) ~ "by chromosome"), y = NULL)+
    scale_color_manual("", values = c("forest green", "dark blue", 'magenta', "red"))+
    
    #add outlier labels
    geom_text(data = EUR.outliers, aes(pop, rate, label = chrom), color = "dark blue", nudge_x = 0.25, size = 2)+
    geom_text(data = AFR.outliers, aes(pop, rate, label = chrom), color = "forest green", nudge_x = 0.25, size = 2)+
    geom_text(data = EAS.outliers, aes(pop, rate, label = chrom), color = "red", nudge_x = 0.25, size = 2)+
    geom_text(data = SAS.outliers, aes(pop, rate, label = chrom), color = "magenta", nudge_x = 0.25, size = 2)+
    
    #add points
    geom_point(aes(color = factor(chrom.dat$pop)), size = .4, position = position_jitter(width = 0.1))+
    theme(axis.text.x = element_text(size = rel(.7)), axis.title.x = element_blank(), #adjust text sizes
          axis.title.y = element_text(size = rel(.7)), axis.text.y = element_text(size = rel(.7)),
          legend.position = 'none',
          title = element_text(size = rel(.5)))#legend position
  return(c_plot)
}

#helper function for ploting which formats the data for ggplot
chrom.process.data <- function(AFR, EUR, EAS, SAS, gw, mut){
  #get indicies for mutation and contextAFR, EUR, EAS, SAS, gw, mut
  i <- which(EUR$Context == mut)
  cntxt <- substr(mut, 1, nchar(mut)-3)
  i.cntxt <- which(gw$Context == cntxt)
  col.e <- ncol(AFR)
  col.s <- col.e-22
  
  #trim summary columns from each input dataframe
  AFR <- AFR[c(col.s: col.e)]
  EUR <- EUR[c(col.s: col.e)]
  EAS <- EAS[c(col.s: col.e)]
  SAS <- SAS[c(col.s: col.e)]
  gw <- gw[-c(1,2)]
  gw_totals <- colSums(gw)
  
  #make output dataframe for plot
  dat <- data.frame(matrix(nrow = 23, ncol = 4))
  colnames(dat) <- c("AFR", "EUR", "SAS", "EAS")  
  
  #get rates for each pop
  dat$'EUR' <- t(EUR[i,]/gw[i.cntxt,]*1.2E-8*gw_totals/colSums(EUR))
  dat$'AFR' <- t(AFR[i,]/gw[i.cntxt,]*1.2E-8*gw_totals/colSums(AFR))
  dat$'EAS' <- t(EAS[i,]/gw[i.cntxt,]*1.2E-8*gw_totals/colSums(EAS))
  dat$'SAS' <- t(SAS[i,]/gw[i.cntxt,]*1.2E-8*gw_totals/colSums(SAS))
  
  #melt data to necessary format
  row.names(dat) <- c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', 'X')
  dat.m <- melt(t(dat))
  colnames(dat.m) <- c("pop", "chrom", "rate")
  
  return(dat.m)
}

#helper function that returns which elements of a vector x are outliers
is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}
```

```{r S4 make figure, fig.height = 8, fig.width=7}
a <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "TCC->T")
b <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "ACC->T")
c <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "TCT->T")
d <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "CCC->T")
e <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "TCG->T")
f <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "ACG->T")
g <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "GCG->T")
h <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "GAT->T")
i <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "ACC->A")
j <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "ACA->T")
k <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "TCA->T")
l <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "ACT->T")
m <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "GCT->T")
n <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "GAC->T")
o <- chrom.box(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, gw_3mer_counts, "GCC->T")


grid.arrange(a,b,c,d,e,f,g,h,i,j,k,l,m,n,o, ncol = 4)
```

**Supplementary Figure 4:** Box plots of inferred private mutation rate across chromosomes for each highly significantly variable 3-mer type.  Outlier points are labeled with chromosome.


***********************************************************************************************************

\pagebreak

```{r S567 upload data}
AFR_7mer_counts <- read_delim("7mer/AFR_7mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
EUR_7mer_counts <- read_delim("7mer/EUR_7mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
EAS_7mer_counts <- read_delim("7mer/EAS_7mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
SAS_7mer_counts <- read_delim("7mer/SAS_7mer_counts.txt", 
                              "\t", escape_double = FALSE, trim_ws = TRUE)
gw_7mer_counts <- read_delim("gw_counts/gw_7mer_counts.txt", 
                             "\t", escape_double = FALSE, trim_ws = TRUE)
source("../R_code/data_wrangling/upload_subpops_7mer.R")
```

```{r S567 functions}
# given: count1 and count2: pop-specific polymorphism dataframes; Mut: 3mer subcontext of interest,
# generates a scatter plot of rate for mut in population 1 by population 2
subrate.scplot <- function(count1, count2, mut, pops = c("Pop 1", "Pop 2"), label = c(), lm = F){
  pop1.mut <- subset(count1, count1$X3mer == mut)
  pop2.mut <- subset(count2, count2$X3mer == mut)
  sample.size <- pop1.mut$Count + pop2.mut$Count
  n <- nchar(mut)
  ref <- substr(mut, 1, n-3)
  alt <- substr(mut, n, n)
  
  #make plot
  p_plot <- qplot(pop1.mut$Rate, pop2.mut$Rate) +
    labs(x = bquote(.(ref)%->%.(alt) ~ .(paste("mutation rate in", pops[1]))), y = bquote(.(ref)%->%.(alt) ~ .(paste("mutation rate in", pops[2])))) +
    theme(legend.title = element_blank(), 
          legend.justification= c(1,0), 
          legend.position=c(1,0),
          legend.text = element_text(size = rel(0.7)),
          legend.key.size = unit(.3, "cm"),
          axis.text=element_text(size = 6),
          text = element_text(size = 7))+
    geom_point(aes(colour = log(base = 10, x = sample.size)), size = .7) + 
    scale_colour_gradient(low = "orange", high = "blue")+
    geom_abline(aes(intercept=0,slope=1), size = .5) + 
    coord_fixed(ratio = 1, 
                xlim = c(0, max(pop1.mut$Rate, pop2.mut$Rate)),
                ylim = c(0, max(pop1.mut$Rate, pop2.mut$Rate)))
  
  if (lm){
    mod1 <- lm(pop2.mut$Rate ~ pop1.mut$Rate)
    pred <- predict(mod1, interval = "prediction")
    
    p_plot <- p_plot + stat_smooth(method = lm, se = FALSE, fullrange = T)+
      geom_line(aes(y = pred[,2]), color = "red", linetype = "dashed", size = 1)+
      geom_line(aes(y = pred[,3]), color = "red", linetype = "dashed", size = 1)
  }
  
  if (length(label!=0)){
    #set aside contexts to label
    dat <- data.frame(cbind(pop1.mut$Context, pop1.mut$Rate, pop2.mut$Rate), stringsAsFactors = F)
    lab <- subset(dat,  is.element(dat[,1], label))
    
    if (length(lab[,1]!=0)){
      lab[,c(2,3)] <- sapply(lab[,c(2,3)], as.numeric)
      p_plot <- p_plot + geom_label_repel(data = lab, aes(lab[,2], lab[,3],
                                   label = lab[,1]), size = 2.5, box.padding = unit(1, "lines"),
                                   point.padding = unit(.3, "lines"),
                                   nudge_x = 1.5e-9, nudge_y = 1.5e-9, color = "purple4")
    }
  }

  return(p_plot)
}
```

```{r S5 make figure, fig.height = 7.5, fig.width=7}
a <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "TCC->T", pops = c("East Asia", "Europe"))
b <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "ACC->T", pops = c("East Asia", "Europe"))
c <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "TCT->T", pops = c("East Asia", "Europe"))
d <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "CCC->T", pops = c("East Asia", "Europe"))
e <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "ACT->T", pops = c("East Asia", "Europe"))
f <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "TCA->T", pops = c("East Asia", "Europe"))
g <- subrate.scplot(EAS_7mer_counts, EUR_7mer_counts, "GCC->T", pops = c("East Asia", "Europe"))
grid.arrange(a,b,c,d,e,f,g, ncol = 3)
```

**Supplementary Figure 5:** Rates of all 7-mer expansions of each profile #1 signature in Europe versus East Asia. Each point represents a 7-mer expansion of the 3-mer subtype shown, plotted based on its estimated mutation rate in each of the two populations displayed. Colors indicate the log (base 10) of the number of substitutions observed for that 7-mer class.  Europe and East Asia were selected to visualize this comparison because the difference in mutation rate for profile #1 3-mers is most strong and consistent between these populations. Generally, these polymorphisms match case II (see main text): enrichment is consistent across 7mers, with some expected amount of noise.

***********************************************************************************************************

\pagebreak

```{r S6 make figure, fig.height = 7.5, fig.width=7}
a <- subrate.scplot(AFR_7mer_counts, EAS_7mer_counts, "GAT->T", pops = c("Africa", "East Asia"))
b <- subrate.scplot(AFR_7mer_counts, EAS_7mer_counts, "ACC->A", pops = c("Africa", "East Asia"))
c <- subrate.scplot(AFR_7mer_counts, EAS_7mer_counts, "GAC->T", pops = c("Africa", "East Asia"))
grid.arrange(a,b,c, ncol = 2)
```

**Supplementary Figure 6:** Rates of all 7-mer expansions of each profile #2 signature in Africa versus East Asia. Again, the population-specific enrichment is consistent across 7mers, consistent with case II.

***********************************************************************************************************

\pagebreak

```{r S7 make figure, fig.height = 7.5, fig.width=7}
a <- subrate.scplot(AFR_7mer_counts, SAS_7mer_counts, "TCG->T", pops = c("Africa", "South Asia"))
b <- subrate.scplot(AFR_7mer_counts, SAS_7mer_counts, "ACG->T", pops = c("Africa", "South Asia"))
c <- subrate.scplot(AFR_7mer_counts, SAS_7mer_counts, "GCG->T", pops = c("Africa", "South Asia"))
d <- subrate.scplot(AFR_7mer_counts, SAS_7mer_counts, "CCG->T", pops = c("Africa", "South Asia"))
grid.arrange(a,b,c, d, ncol = 2)
```

**Supplementary Figure 7:** Rates of all 7-mer expansions of each profile #4 signature in Africa versus South Asia. The relative mutation rate variability between the CpG mutations is small compared to other profiles noted in this report. However, we still observe that most 7-mer expansions of these CpG substitutions lie above the diagonal, consistent with case II.

***********************************************************************************************************

\pagebreak

```{r S8 functions}
# calculates homogeneity test p values for pairwise comparisons of two dfs of counts
pairwise.chi <- function(counts.1, counts.2, filter = T){
  n.contexts = length(counts.1$Context)
  result <- data.frame(matrix(ncol=2,nrow=n.contexts))
  colnames(result) <- c("Context", "p")
  
  result$Context <- counts.1$Context
  sum.1 <- sum(counts.1$Count)
  sum.2 <- sum(counts.2$Count)

  for (i in 1:n.contexts){
    c.a <- c(counts.1$Count[i], counts.2$Count[i])
    c.b <- c(sum.1, sum.2) - c.a
    data <- cbind(c.a, c.b)
    warning <- is(tryCatch(chisq.test(data), warning = function(w) w), "warning")
    if (filter == T & warning){
      result$p[i] <- NA
    }
    else result$p[i] <- chisq.test(data)$p.value
  }
  
  return(result)
}
```


```{r S8 make figure, fig.height = 7.5, fig.width=7}
p.CDXJPT <- pairwise.chi(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts)
p.CDXJPT <- subset(p.CDXJPT, is.element(CDX_EAS_7mer_counts$X3mer, c("AAC->C", "CAC->C", "TAT->T", "GAC->C", "TAC->C")))
p.CDXJPT$fdr <- p.adjust(p.CDXJPT$p, method = "fdr")

JPT7mers <- subset(p.CDXJPT$Context, p.CDXJPT$fdr < 0.05)

a <- subrate.scplot(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts, "AAC->C", pops = c("Chinese Dai", "Japanese"), label = JPT7mers)
b <- subrate.scplot(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts, "CAC->C", pops = c("Chinese Dai", "Japanese"), label = JPT7mers)
c <- subrate.scplot(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts, "GAC->C", pops = c("Chinese Dai", "Japanese"), label = JPT7mers)
d <- subrate.scplot(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts, "TAC->C", pops = c("Chinese Dai", "Japanese"), label = JPT7mers)
e <- subrate.scplot(CDX_EAS_7mer_counts, JPT_EAS_7mer_counts, "TAT->T", pops = c("Chinese Dai", "Japanese"), label = JPT7mers)
grid.arrange(a,b,c,d,e, ncol = 2)
```

**Supplementary Figure 8:** Rates of all 7-mer expansions of each profile #3 signature in Japanese versus Chinese Dai. Labeled points are those which were significantly variable between Japanese and Chinese Dai (fdr < 0.05 among all profile #3 7-mer expansions).

***********************************************************************************************************

\pagebreak 

```{r S9 functions}
CI.plot.subpop <- function(subpops, mut, popnames, colors = colors.ancestral){
  n <- nchar(mut)
  ref <- substr(mut, 1, n-3)
  alt <- substr(mut, n, n)
  n.pops <- length(subpops)
  mut.i <- which(subpops[[1]]$Context == mut)
  
  #have to do a silly workaround or R will sort popnames alphanumerically
  poplabs <- factor(popnames, levels= popnames)
  
  #cycle through subpops and get counts for mut
  counts <- rep(0, n.pops)
  sums <- rep(0, n.pops)
  for (i in 1:n.pops){
    counts[i] <- subpops[[i]]$Count[mut.i]
    sums[i]<- sum(subpops[[i]]$Count)
  }
  
  #estimate substitution probability
  N.c <- subpops[[1]]$context_in_genome[mut.i] 
  theta <- counts/N.c
  L <- theta - 1.96*sqrt(theta*(1-theta)/N.c)
  U <- theta + 1.96*sqrt(theta*(1-theta)/N.c)
  
  #normalize to rate estimate; assume genome wide subsitution probability is measured without error
  norm <- 1.2E-8*(sum(as.numeric(subpops[[1]]$context_in_genome))/3)/sums
  
  
  df <- data.frame(cbind(poplabs, theta*norm, L*norm, U*norm))
  plotcol <- reorder(colors, theta*norm)
  
  ggplot(df, aes(reorder(popnames, poplabs), theta*norm)) +
    geom_point(size = 4, color = plotcol) +
    geom_errorbar(aes(ymax = U*norm, ymin = L*norm), color = plotcol, size = 0.75)+
    labs(y = bquote("Estimated mutation rate of" ~ .(ref)%->%.(alt))) + #y axis label
    theme(axis.text.x = element_text(size = rel(1), angle = 45, hjust = 1), axis.title.x = element_blank(), #adjust text sizes
          axis.title.y = element_text(size = rel(1)), axis.text.y = element_text(size = rel(.9)),
          legend.position = 'none')
}
```

```{r}
EASsubpops <- list(CDX_EAS_7mer_counts, KHV_EAS_7mer_counts, CHS_EAS_7mer_counts, CHB_EAS_7mer_counts, JPT_EAS_7mer_counts)
EASpopnames <- c("CDX", "KHV", "CHS", "CHB", "JPT")
CI.plot.subpop(EASsubpops, "AAACAAA->A", EASpopnames, rep("red", 5))
```

**Supplementary Figure 9:** Approximate 95% confidence interval estimates of inferred mutation rate for AAACAAA$\rightarrow$A across East Asian subpopulations: Chinese Dai in Xishuangbanna (CDX); Kinh in Ho Chi Minh City, Vietnam (KHV); Southern Han Chinese (CHS); Han Chinese in Beijing (CHB); and Japanese in Tokyo.  This pattern resembles that of 3-mer profile #3.

***********************************************************************************************************

\pagebreak 

```{r S table 123 data}
AFR_5mer_counts <- read.delim('5mer/AFR_5mer_counts.txt')
EUR_5mer_counts <- read.delim('5mer/EUR_5mer_counts.txt')
EAS_5mer_counts <- read.delim('5mer/EAS_5mer_counts.txt')
SAS_5mer_counts <- read.delim('5mer/SAS_5mer_counts.txt')
```

```{r S table 123 functions}
# calculates homogeneity test p values for Fourway comparisons of counts dfs
fourway.chi <- function(AFR, EUR, EAS, SAS, filter = T){
  n.contexts = length(AFR$Context)
  
  # make dataframe for results
  result <- data.frame(matrix(ncol=9,nrow=n.contexts))
  colnames(result) <- c("Context", "X5mer","X3mer", "X1mer", 
                        "AFR.Count", "EUR.Count", "EAS.Count", "SAS.Count", "p")
  result$Context <- AFR$Context
  result$X5mer <- AFR$X5mer # for smaller contexts, X3mer and X5mer columns do not exist,
  result$X3mer <- AFR$X3mer # and will disappear at this step
  result$X1mer <- AFR$X1mer 
  result$AFR.Count <- AFR$Count; result$EUR.Count <- EUR$Count
  result$EAS.Count <- EAS$Count; result$SAS.Count <- SAS$Count
  
  # start setting up tables
  sums <- c(sum(AFR$Count), sum(EUR$Count), sum(EAS$Count), sum(SAS$Count))
  
  # set up table and run test for each context
  for (i in 1:n.contexts){
    c.a <- c(AFR$Count[i], EUR$Count[i], EAS$Count[i], SAS$Count[i])
    c.b <- sums - c.a
    data <- cbind(c.a, c.b)
    warning <- is(tryCatch(chisq.test(data), warning = function(w) w), "warning")
    if (filter == T & warning){
      result$p[i] <- NA}
    else result$p[i] <- chisq.test(data)$p.value
    }
  return(result)
}
ordered.p <- function(pdata){
  #preprocess data to order and remove nas 
  pdata <- pdata[complete.cases(pdata$p),]
  myorder <- order(pdata$p)
  n.muts <- length(pdata$p)
  
  p.ordered <- rep(0, n.muts)
  
  #set largest p-value 
  j <- myorder[n.muts]
  p.ordered[j] <- pdata$p[j]
  
  #initialize not mutated counts based on this lowest p-value mutation
  not.mut <- c(pdata$AFR.Count[j], pdata$EUR.Count[j], 
               pdata$EAS.Count[j], pdata$SAS.Count[j])
  
  for (i in n.muts:1){
    j <- myorder[i]
    mut <-c(pdata$AFR.Count[j], pdata$EUR.Count[j], 
            pdata$EAS.Count[j], pdata$SAS.Count[j])
    data <- cbind(mut, not.mut)
    p.ordered[j] <- chisq.test(data)$p.value
    
    #add these mutations to the not.mutated counts for future tests
    not.mut <- not.mut + mut
  }
  pdata$p <- p.ordered
  return(pdata)
}
```


```{r S table 1 make}
# calculate ordered p values
p.3mer <- fourway.chi(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts)
o.p.3mer <- ordered.p(p.3mer)
o.p.3mer <- o.p.3mer[order(o.p.3mer$p),]
o.sig.3mers <- subset(o.p.3mer, o.p.3mer$p < 0.05/96)
colnames(o.sig.3mers) <- c("Context", "1-mer", "Counts in AFR", "Counts in EUR", "Counts in EAS", "Counts in SAS", "p")

# save
write.table(o.sig.3mers, "supplemetary_table_1.csv", quote = F, sep = ",", row.names = F)
```

```{r table S2 make}
# calculate ordered p values
p.5mer <- fourway.chi(AFR_5mer_counts, EUR_5mer_counts, EAS_5mer_counts, SAS_5mer_counts)
o.p.5mer <- ordered.p(p.5mer)
o.p.5mer <- o.p.5mer[order(o.p.5mer$p),]
o.sig.5mers <- subset(o.p.5mer, o.p.5mer$p < 0.05/(dim(o.p.5mer)[1]))
colnames(o.sig.5mers) <- c("Context", "3-mer", "1-mer", "Counts in AFR", "Counts in EUR", "Counts in EAS", "Counts in SAS", "p")

# save
write.table(o.sig.5mers, "supplemetary_table_2.csv", quote = F, sep = ",", row.names = F)
```

```{r table S3 make}
# calculate ordered p values
p.7mer <- fourway.chi(AFR_7mer_counts, EUR_7mer_counts, EAS_7mer_counts, SAS_7mer_counts)
o.p.7mer <- ordered.p(p.7mer)
o.p.7mer <- o.p.7mer[order(o.p.7mer$p),]
o.sig.7mers <- subset(o.p.7mer, o.p.7mer$p < 0.05/(dim(o.p.7mer)[1]))
colnames(o.sig.7mers) <- c("Context","5-mer", "3-mer", "1-mer", "Counts in AFR", "Counts in EUR", "Counts in EAS", "Counts in SAS", "p")

# save
write.table(o.sig.7mers, "supplemetary_table_3.csv", quote = F, sep = ",", row.names = F)
```
