---
title: "5mer Datagazing"
author: "Rachael 'Rocky' Aikens, Voight Lab"
date: "July 10, 2017"
output: html_document
---

Need to source CI_and_chrom_plots_report.Rmd before this one.

# 5mer signals of mutation rate differences

Recall the following list of top variable 5mers, with known 3mer signals (GGA->A, ACC->T, AGA->A, ACT->A, ACC->S, and ACA->T) removed:

```{r print sig.5mers, echo = F}
kable(head(sig.5mers, n = 10), row.names = F, digits = 600,
      caption = "10 most highly significant 5mers")
```


##TTAAA->T 

The most significant novel signal of rate variation at the 5mer level is TTAAA->T, which appears elevated in Africa.  Interestingly, this polymorphism type has a much higher rate of occurance than other 3mers with the TAA->A subcontext. As we will see, this signal may be highly variable at a 7mer level.

##GGCCC->T

one will note that the 3mer subcontext of this mutation type appears as highly significantly variable in our original heterogeneity test.  However, we can see that this polymorphism varies in a different way than the other 5mers with the same threemer subtype.  GGCCC->T more closely resembles the profile of ACT->A and ACC->A.

```{r}
a <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GCC->T")
b <- CI.plot(AFR_5mer_counts, EUR_5mer_counts, EAS_5mer_counts, SAS_5mer_counts, "GGCCC->T")
c <- subrate.lplot(AFR_5mer_counts, EUR_5mer_counts, EAS_5mer_counts, SAS_5mer_counts, "GCC->T")
grid.arrange(a,b, ncol = 2)
c
```
```{r}
a <- CI.plot(AFR_3mer_counts, EUR_3mer_counts, EAS_3mer_counts, SAS_3mer_counts, "GCT->T")
b <- CI.plot(AFR_5mer_counts, EUR_5mer_counts, EAS_5mer_counts, SAS_5mer_counts, "GGCTT->T")
c <- subrate.lplot(AFR_5mer_counts, EUR_5mer_counts, EAS_5mer_counts, SAS_5mer_counts, "GCT->T")
grid.arrange(a,b, ncol = 2)
c
```

Two 7mer expansions of TTAAA->T, "TTTAAA->T" and "ATTAAAA->T" show a very different pattern of variation than the other TAA->T polymorphisms:

```{r, echo = F, warning = F}
subrate.scplot(SAS_5mer_counts, AFR_5mer_counts, "TAA->T")
```